Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-24 src/1.Introduction.tex
Fused DZP and JC changes for Chapters 1 and 2.

Parallel and distributed computing moved into focus at universities and teaching institutions. The interest is two-fold. Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula, and secondly, parallel computing is an attractive field for department research and industry collaborations. At Sandia, we develop a performance portability library called Kokkos that is ideal for educators looking to teach parallel computing.

In teaching a class on parallel programming, educators must balance considerations of architecture, programming model, and theoretical concepts. Unfortunately, not all architectures support all programming models, nor do many programming models cleanly express all parallel programming constructs. Many vendor models compound this, requiring students have access to specific hardware, and limiting the concepts of parallelism to those supported by that architecture. This makes the choice of programming model important and difficult.

Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency. In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine model specification. We would like to compare this to the support of Von Neumann architectures in programming languages today. We define an abstract machine model later in this work. The lack of such parallel programming models motivated us to develop Kokkos.

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL) for developing million line scientific software on the world\'s biggest supercomputers. It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes functions representing parallel patterns and types for data management and maps them to the underlying abstract execution model. This execution model creates a generic view of compute hardware and memories. The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}. Its key properties are

@@

Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards architecture-independent and performance-portable programming of CPUs and accelerators. We show an abstraction with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

Secondly, we would like to promote Kokkos to students and educators as a platform to understand and experiment with parallel patterns in a fashion that is

\begin{itemize}
\item Generic
\item Vendor-agnostic
\item Declarative
\end{itemize}

First, Kokkos has semantics that are generic. This means that students are learning parallelism detached from any architectural concerns. Kokkos won\'t require them to know the specifics of the \"symmetric multiprocessors\" of a GPU or the cache hiearchy on a CPU\footnote{Though nothing precludes an educator from teaching these independently of the model}. Finally, Kokkos is vendor agnostic. If you have a computer lab with processors from one vendor and your students have processors from another, this presents no problem, Kokkos likely works on both. Kokkos is declarative, students aren\'t saying \"map these work items to this set of threads\" but \"do a reduction over this set of data.\" This is critical, much of software engineering is moving to declarative models and languages, learning to operate in such a mindset is useful outside of parallel programming.

>>>>

Parallel and distributed computing is a focus at universities and teaching institutions. The interest is two-fold. Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula, and secondly, parallel computing is an attractive field for department research and industry collaborations.

While industry-driven collaborations often predefine technological frameworks such as programming environments and targeted architectures, academic teaching leaves this mostly undefined. In this case, educators are exposed to a variety of solutions, approaches and trends in parallel computing. Which particular set to choose for teaching in class is not trivial. While decision factors differ, maximizing the subject relevance by teaching generic principles of parallel programming generically is among the most important ones.

Teaching generic parallel programming today is difficult. Educators strive to balance considerations of architecture, programming model, and theoretical concepts. Unfortunately, not all architectures support all programming models, nor do programming models offer enough abstractions to support modern hardware architectures sufficiently . Many vendor programming models compound this, requiring students to have access to specific hardware, and limiting the concepts of parallelism to those supported by that hardware.

Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency. In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine specification. We would like to compare this to the support of Von Neumann architectures in programming languages today. We define an abstract machine model later in this work. The lack of such parallel programming models motivated us to develop Kokkos.

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL). It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}. We list its key properties as follows.

@@

Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards architecture-independent and performance-portable programming of CPUs and accelerators. We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

Secondly, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative. This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.

### 2020-01-24 src/2.GenericProgramming.tex
Fused DZP and JC changes for Chapters 1 and 2.

A programming model exposes constrained semantics to the programmer which that programmer uses to express intent. Absent an embedded programming model, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance portable code. This is commonly achieved by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others. Some representatives are the OpenMP programming model~\cite{CITEOPENMP}, threading libraries, or languages like Erlang in which parallelism is a first-class citizen. In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they combine convenience to the programmer with semantics sufficient to map the program execution to the parallel computer hardware efficiently and correctly.

In designing Kokkos with this mentality, we looked at a broad variety of current and future architectures on which software would need to run. We also looked at our codes, which express many different algorithmic property. We conceived of Kokkos as a generic programming system that could cover the space of architectures we cared about, while exposing semantics in which the algorithms of our codes could be easily expressed. In this way, a generic programming system embedded in the language became the glue between the semantics of a problem domain and the capabilities of an architecture domain.

\begin{figure}
\centerline{\includegraphics[width=0.45\textwidth]{img/Hollman.png}}
\caption{The Hollman pic :) - placeholder}
\label{fig}
\end{figure}

Earlier we discuss how a programming model constrains semantics to allow the expression of intent in such a way that a developer achieves their goals. Putting together an effective programming model requires constraining semantics such that they enable the necessary parallel paradigms, and expose to users only those aspects of the underlying hardware that they need to think about. We call the combination of constrained semantics, parallel paradigms, and division of responsibility between developer and framework the Semantic Capture.

\begin{figure}[h]
\begin{Verbatim}[frame=leftline]
Constrained Semantics
Parallel Paradigms 
Division of Responsibility
\end{Verbatim}
\caption{Semantic capture - a triplet of constrained semantics, parallel paradigms, and division of responsibilities defined by a parallel programming model to allow developers to run performance-portable code on a set of architectures}
\label{figSemCapture}
\end{figure}

%Semantic information provided by the programmer to the parallel programming model can be grouped into information to express intent and properties. They address the question of \emph{what}, \emph{where} and \emph{how}. In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run and access and how to run that parallel code. Synchronization primitives may be considered as part of the what and information on the execution properties such as data placement or memory access type as the how. Semantics are expressed following a paradigm and a programming language.

To enable developers to achieve performance portability, we needed to constrain what intent they could specify to Kokkos. We settled on Kokkos supporting semantics for offloading loops, the allocation and freeing of data, and data movement. Critically, we decided against exposing machine details to the developer, we aim to increase portability of code by only exposing an abstract execution model (described later) to the users. Our semantics map naturally to C-like languages, with malloc/new calls replaced with the construction and destruction of abstract data types, and many loop types replaced with corresponding parallel patterns. This clean mapping between traditional C-like semantics and our model came from requirements from our physics teams, but also makes Kokkos a good choice for students as it will more closely track with what they've seen before.

%TODO DZP: keep going here

@@

Lastly, a model must decide between declarative and imperative semantics. Declarative semantics require a developer to express intent, implementation is the responsibility of the programming model and underlying toolchain. Imperative semantics require the developer to specify the mechanisms by which their intent will be implemented. In an imperative model, developer success is tied to their knowledge of implementation mechanisms. In a declarative model that success is tied to the developer's ability to accurately declare their requirements, and the toolchain's ability to turn that into an implementation. Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned types of semantics. While both programming models implement the same programming paradigm (parallel patterns), Figure~\ref{figOMPLike} required the developer to express how to parallelize the loop, while in~\ref{figKokkosLike} the developer just declared their parallel semantics, and relied on the programming model to pick an implementation.

Kokkos takes the declarative approach. For the developers of our large scientific applications, this is appealing because it allows them to express their algorithms in one way, and rely on Kokkos to map that intent effectively to a variety of architectures. For students this is appealing because they're unlikely to have expertise in these architectures. For an educator looking for an implementation vehicle with which to teach students about parallelism, the declarative model eliminates the burden of reimplementing exercises as computer labs and dominant architectures change. For a declarative model to make sense, the educator should know something about the abstract execution model being used to ensure performance portability. Depending on the nature of the course, this could be mostly hidden from students, or discussed in detail.

\begin{figure}
\begin{Verbatim}[frame=leftline]
# pragma model parallel for simd vectorize tile unroll fuse
for ( size_t i = 0; i < N; ++i) {
 /* loop body */
}
\end{Verbatim}
\caption{Imperative model tells the compiler to use simd and vectorize and tile and unroll and fuse. It lists specific optimizations that should be made}
\label{figOMPLike}
\end{figure}

\begin{figure}
\begin{Verbatim}[frame=leftline]
model::parallel_for (N, [=] ( const size_t i) {
  /* loop body */
});
\end{Verbatim}
\caption{Declarative model expresses parallelism and relies on the model to map it effectively to the architecture}
\label{figKokkosLike}
\end{figure}

>>>>

A programming model exposes constrained semantics to the programmer which the programmer uses to express intent. Absent an embedded programming model, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance portable code. This is commonly achieved by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others. Some representatives are the OpenMP programming model~\cite{CITEOPENMP}, threading libraries, or languages like Erlang in which parallelism is a first-class citizen. In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they combine convenience to the programmer with semantics sufficient to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present and by what syntactical means span an exciting design space worth exploring. From a theoritecal perspective

 Figure~\ref{figSemCapture} shows the three  areas that constitute a parallel programming model and its design. We call it the semantic capture.

\begin{figure}[h]
\begin{Verbatim}[frame=leftline]
Constrained Semantics: memory allocations, memory movement, parallel kernels
Parallel Paradigms: patterns, tasks
Expression: language 
Concern: prescriptive, descriptive
\end{Verbatim}
\caption{Semantic capture - a triplet of constrained semantics, parallel paradigms, and expression define by a parallel programming model to allow developers to run performance-portable code on a set of architectures}
\label{figSemCapture}
\end{figure}

Constrained semantics provided by the programmer to the parallel programming model can be grouped into information to express intent and properties. They address the question of \emph{what}, \emph{where} and \emph{how}. In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run and access and how to run that parallel code. Synchronization primitives may be considered as part of the what and information on the execution properties such as data placement or memory access type as the how. Semantics are expressed following a paradigm and a programming language.

@@

Lastly, a language defines the syntax of the parallel semantic. Annotations-based  languages add semantic information to the base language through pragmas while embedded programming models rely on the base language and offer additional programming interfaces to capture information. Their semantic is defined in the API specification.

Lastly, a programming model must divide responsibility between the developer and the model. We refer to is as defining the concern. In practice, this is a choice between defining a descriptive and prescriptive semantics. Descriptive semantics require a developer to express intent. The implementation is the responsibility of the programming model and of the underlying toolchain. Prescriptive semantics require the developer to specify the mechanisms by which their intent will be implemented. In an prescriptive model, developer success is tied to their knowledge of implementation mechanisms. In a descriptive model that success is tied to the developer's ability to accurately describe their requirements and the toolchain's ability to turn that into an implementation.

Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned parallel programming syntaxes. While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations. Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} can have a descriptive or prescriptive semantic of describing a parallel loop construct. It is up to the programming model specification to make further definitions regarding separation of concern.

Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow to map parallel execution on modern architectures. In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first. In the following chapters we use the words \emph{representation} and \emph{model} interchangeably. In this work the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.

\begin{figure}
\begin{Verbatim}[frame=leftline]
# pragma model parallel loop
for ( size_t i = 0; i < N; ++i) {
 /* loop body */
}
\end{Verbatim}
\caption{Annotations-based programming models extend the semantic of the base language through pragma annotations.}
\label{figOMPLike}
\end{figure}

\begin{figure}
\begin{Verbatim}[frame=leftline]
model::parallel_for (N, [=] ( const size_t i) {
  /* loop body */
});
\end{Verbatim}
\caption{Embedded parallel programming uses the base language and relies on programming interfaces to capture information.}
\label{figKokkosLike}
\end{figure}

### 2020-01-24 src/3.MachineModel.tex
Further renaming.
Fused DZP and JC changes for Chapters 1 and 2.

\section{The Kokkos Execution Model Abstractions}\label{chap:kokkosEMA}

@@

The Kokkos abstract machine model targets a design of a future shared-memory computing architectures. The design is shown in Figure~\ref{fig:execspace}. It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators. Both compute device types can have disjoint memory address spaces with unique performance properties. In such an architecture, execution units might be hierarchically organized with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches. In order to ensure performance portability on such a wide range of configurations, an abstraction of compute resources and available memories are required. For this purpose we introduce the abstract concept called \emph{space}.

>>>>

\section{The Kokkos Machine Model}\label{chap:kokkosMM}

@@

The Kokkos abstract machine representation targets a design of a future shared-memory computing architectures. The design is shown in Figure~\ref{fig:execspace}. It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators. Both compute device types can have disjoint memory address spaces with unique performance properties. In such an architecture, execution units might be hierarchically organized with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches. In order to ensure performance portability on such a wide range of configurations, an abstraction of compute resources and available memories are required. For this purpose we introduce the abstract concept called \emph{space}.
